---
title: "RNA-seq Analysis"
author: "Bruna Toledo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This notebook integrates the main RNA-seq analysis steps:

1. Removal of unwanted variation with **RUVSeq**
2. Differential expression analysis with **DESeq2**
3. QC, visualization and functional enrichment analysis
4. Heatmap visualization for selected genes from GSEA-enriched signatures

This notebook mirrors the standalone scripts:

- `01_remove_unwanted_variation.R`
- `02_differential_expression.R`
- `03_deg_visualization_and_enrichment.R`
- `04_GSEA_heatmap.R`

# 1. Load libraries

```{r load-libraries}
suppressPackageStartupMessages({
  library(RUVSeq)
  library(EDASeq)
  library(edgeR)
  library(RColorBrewer)
  library(readr)
  library(dplyr)
  library(ashr)
  library(DESeq2)
  library(writexl)
  library(stringr)
  library(rtracklayer)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(ggrepel)
  library(forcats)
  library(openxlsx)
  library(pheatmap)
  library(ReactomePA)
  library(enrichplot)
  library(clusterProfiler)
  library(EnhancedVolcano)
  library(tibble)
  library(matrixStats)
  library(ggplotify)
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
})
```

# 2. Define input and output paths

```{r define-paths}
salmon_counts <- "salmon.merged.gene_counts.tsv"
samples_tsv   <- "samplesheet.tsv"
tx2gene_file  <- "tx2gene.tsv"
gtf_file      <- "gencode.v49.primary_assembly.annotation.gtf"

ruv_dir          <- "1_RUVseq_output"
deg_tables_dir   <- "1_DEG_tables"
qc_dir           <- "2_QC_output"
deg_overview_dir <- "3_DEG_overview"
enrich_dir       <- "4_DEG_Enrichment"
tf_dir           <- "5_TFs"
gsea_dir         <- "6_GSEA"

dir.create(ruv_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(deg_tables_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(deg_overview_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(enrich_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(tf_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(gsea_dir, showWarnings = FALSE, recursive = TRUE)

out_base   <- "salmon.merged.gene_counts_RUVseq"
out_prefix <- file.path(ruv_dir, out_base)
```

# 3. Remove unwanted variation (RUVSeq)

## 3.1 Load counts and metadata

```{r load-counts-metadata}
counts_raw <- read_tsv(salmon_counts, show_col_types = FALSE) %>%
  as.data.frame()

rownames(counts_raw) <- counts_raw$gene_id
counts_raw <- counts_raw[, 3:14]
counts_raw <- as.matrix(counts_raw)

metadata <- read_tsv(samples_tsv, show_col_types = FALSE) %>%
  mutate(
    subject   = factor(subject),
    condition = factor(condition),
    batch     = as.character(batch)
  )

sample_ids <- as.character(metadata$sample)

stopifnot(setequal(colnames(counts_raw), sample_ids))
counts_raw <- counts_raw[, sample_ids]
stopifnot(identical(colnames(counts_raw), sample_ids))
```

## 3.2 Filter low-abundance genes

```{r filter-counts}
keep_abundance <- rowSums(counts_raw >= 25, na.rm = TRUE) >= 3
keep_nonNA     <- rowSums(!is.na(counts_raw)) >= (ncol(counts_raw) * 0.5)
keep_both      <- keep_abundance & keep_nonNA

counts_ftd <- counts_raw[keep_both, , drop = FALSE]
counts_ftd <- round(counts_ftd)
```

## 3.3 Create `SeqExpressionSet`

```{r make-seqexpressionset}
set_raw <- newSeqExpressionSet(
  counts = counts_ftd,
  phenoData = data.frame(
    row.names = metadata$sample,
    condition = metadata$condition,
    subject   = metadata$subject,
    batch     = metadata$batch
  )
)
```

## 3.4 PCA before and after normalization

```{r pca-raw-uq}
colors <- brewer.pal(n = nlevels(metadata$condition), name = "Set2")

plotPCA(set_raw, col = colors[metadata$condition], cex = 1.2)

set_uq <- betweenLaneNormalization(set_raw, which = "upper")

plotPCA(set_uq, col = colors[metadata$condition], cex = 1.2)
```

## 3.5 Estimate unwanted variation factors with RUVr

```{r ruvr}
design <- model.matrix(~ condition, data = pData(set_uq))

y <- DGEList(
  counts = EDASeq::counts(set_uq),
  group  = metadata$condition
)

y <- calcNormFactors(y, method = "upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)
fit <- glmFit(y, design)

resids <- residuals(fit, type = "deviance")

k <- 2
controls <- rownames(EDASeq::counts(set_uq))

set_ruvr <- RUVr(
  set_uq,
  controls,
  k = k,
  res = resids
)

pData(set_ruvr)
```

```{r pca-ruvr}
plotPCA(set_ruvr, col = colors[metadata$condition], cex = 1.2)
```

# 4. Differential expression analysis (DESeq2)

## 4.1 Helper functions

```{r deseq2-helpers}
get_results <- function(dds, contrast, alpha = 0.05, shrink = TRUE) {
  res <- DESeq2::results(dds, contrast = contrast, alpha = alpha)

  if (shrink) {
    if (!requireNamespace("ashr", quietly = TRUE)) {
      stop("Package 'ashr' is required for LFC shrinkage.")
    }
    res <- DESeq2::lfcShrink(dds, contrast = contrast, type = "ashr")
  }

  out <- as.data.frame(res)
  out$gene_id <- rownames(out)
  out <- out[, c("gene_id", setdiff(names(out), "gene_id"))]
  out
}

filter_deg <- function(df,
                       lfc_col = "log2FoldChange",
                       padj_col = "padj",
                       alpha = 0.05,
                       lfc_thr = log2(1.5)) {
  stopifnot(all(c(lfc_col, padj_col) %in% colnames(df)))

  df %>%
    mutate(regulation = ifelse(.data[[lfc_col]] > 0, "Up", "Down")) %>%
    filter(!is.na(.data[[padj_col]])) %>%
    filter(.data[[padj_col]] < alpha, abs(.data[[lfc_col]]) >= lfc_thr)
}
```

## 4.2 Prepare DESeq2 design

```{r prepare-deseq2}
cd <- as.data.frame(pData(set_ruvr), stringsAsFactors = FALSE)

cd$subject   <- factor(cd$subject)
cd$condition <- factor(cd$condition)
cd$condition <- relevel(cd$condition, ref = "Pre-Abx")

use_W <- c("W_2")
rhs <- c("subject", use_W, "condition")
design_formula <- as.formula(paste("~", paste(rhs, collapse = " + ")))

cts <- round(EDASeq::counts(set_ruvr))
stopifnot(identical(colnames(cts), rownames(cd)))

dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData   = cd,
  design    = design_formula
)

dds <- DESeq2::DESeq(dds, test = "Wald", fitType = "parametric")
```

## 4.3 Define contrasts and compute DEGs

```{r run-deseq2}
contrasts <- list(
  PostAbx_vs_PreAbx       = c("condition", "Post-Abx",     "Pre-Abx"),
  PreAbxLPS_vs_PreAbx     = c("condition", "Pre-Abx+LPS",  "Pre-Abx"),
  PostAbxLPS_vs_PostAbx   = c("condition", "Post-Abx+LPS", "Post-Abx"),
  PostAbxLPS_vs_PreAbxLPS = c("condition", "Post-Abx+LPS", "Pre-Abx+LPS"),
  PostAbx_vs_PreAbxLPS    = c("condition", "Post-Abx",     "Pre-Abx+LPS")
)

DEG_res_list <- lapply(
  contrasts,
  function(ct) get_results(dds, ct, alpha = 0.05, shrink = TRUE)
)
```

## 4.4 Annotate DEG tables

```{r annotate-degs}
gene_anno <- read_tsv(tx2gene_file, show_col_types = FALSE) %>%
  as.data.frame() %>%
  mutate(
    transcript_id = as.character(transcript_id),
    gene_id       = as.character(gene_id),
    gene_name     = as.character(gene_name)
  )

gtf <- import(gtf_file)

gene_annot <- gtf %>%
  as.data.frame() %>%
  filter(type == "gene") %>%
  transmute(
    gene_id   = gene_id,
    gene_name = gene_name,
    gene_type = gene_type
  ) %>%
  distinct()

gene_map <- setNames(
  as.character(gene_annot$gene_name),
  as.character(gene_annot$gene_id)
)

gene_type_map <- setNames(
  gene_annot$gene_type,
  gene_annot$gene_id
)

DEG_res_list <- lapply(DEG_res_list, function(df) {
  stopifnot("gene_id" %in% names(df))
  df$gene_name <- gene_map[as.character(df$gene_id)]
  df$gene_type <- gene_type_map[as.character(df$gene_id)]
  df
})

deg_list_sig <- lapply(DEG_res_list, function(df) {
  filter_deg(df, alpha = 0.05, lfc_thr = log2(1.5))
})

head(DEG_res_list[[1]])
```

## 4.5 Normalized counts

```{r normalized-counts}
norm_counts <- as.data.frame(set_ruvr@assayData[["normalizedCounts"]])

norm_counts_anno <- norm_counts
norm_counts_anno$gene_id   <- rownames(norm_counts)
norm_counts_anno$gene_name <- gene_map[norm_counts_anno$gene_id]

norm_counts_anno <- norm_counts_anno[, c(
  "gene_id",
  "gene_name",
  setdiff(names(norm_counts_anno), c("gene_id", "gene_name"))
)]

head(norm_counts_anno[, 1:6])
```

# 5. QC and visualization

## 5.1 Helper functions

```{r plot-helpers}
stop_if_non_numeric <- function(x) {
  if (!is.matrix(x)) x <- as.matrix(x)
  if (!is.numeric(x)) storage.mode(x) <- "numeric"
  x
}

log2p1 <- function(m) log2(m + 1)

map_aes <- function(meta) {
  aes_list <- list(color = NULL, shape = NULL)
  if ("condition" %in% names(meta)) aes_list$color <- meta$condition
  if ("batch" %in% names(meta))     aes_list$shape <- meta$batch
  aes_list
}

do_pca_plot <- function(m, meta, title) {
  m <- stop_if_non_numeric(m)
  X <- t(m)
  keep <- apply(X, 2, sd, na.rm = TRUE) > 0
  X <- X[, keep, drop = FALSE]
  pc <- prcomp(X, center = TRUE, scale. = FALSE)
  pc_df <- as.data.frame(pc$x[, 1:2, drop = FALSE])
  pc_df$sample <- rownames(pc$x)
  df <- left_join(pc_df, meta, by = c("sample" = "sample"))

  var_exp <- (pc$sdev^2) / sum(pc$sdev^2)
  xlab <- sprintf("PC1 (%.1f%%)", 100 * var_exp[1])
  ylab <- sprintf("PC2 (%.1f%%)", 100 * var_exp[2])

  aes_map <- map_aes(df)
  p <- ggplot(df, aes(PC1, PC2)) +
    {
      if (!is.null(aes_map$color) && !is.null(aes_map$shape)) {
        geom_point(aes(color = aes_map$color, shape = aes_map$shape), size = 3)
      } else if (!is.null(aes_map$color)) {
        geom_point(aes(color = aes_map$color), size = 3)
      } else {
        geom_point(size = 3)
      }
    } +
    labs(title = title, x = xlab, y = ylab, color = "condition", shape = "batch") +
    theme_bw()
  p
}

do_mds_plot <- function(m, meta, title) {
  m <- stop_if_non_numeric(m)
  X <- t(m)
  keep <- apply(X, 2, sd, na.rm = TRUE) > 0
  X <- X[, keep, drop = FALSE]
  d <- dist(scale(X))
  mds <- cmdscale(d, k = 2)
  df <- data.frame(MDS1 = mds[, 1], MDS2 = mds[, 2], sample = rownames(mds))
  df <- left_join(df, meta, by = "sample")

  aes_map <- map_aes(df)
  p <- ggplot(df, aes(MDS1, MDS2)) +
    {
      if (!is.null(aes_map$color) && !is.null(aes_map$shape)) {
        geom_point(aes(color = aes_map$color, shape = aes_map$shape), size = 3)
      } else if (!is.null(aes_map$color)) {
        geom_point(aes(color = aes_map$color), size = 3)
      } else {
        geom_point(size = 3)
      }
    } +
    labs(title = title, color = "condition", shape = "batch") +
    theme_bw()
  p
}

long_expr <- function(m, meta) {
  m <- stop_if_non_numeric(m)
  df <- as.data.frame(m, check.names = FALSE)
  df$gene <- rownames(df)
  df <- pivot_longer(df, -gene, names_to = "sample", values_to = "expr")
  left_join(df, meta, by = "sample")
}
```

## 5.2 Library sizes

```{r library-sizes}
counts_raw  <- stop_if_non_numeric(counts_raw)
counts_ftd  <- stop_if_non_numeric(counts_ftd)
norm_counts <- stop_if_non_numeric(norm_counts)

libsizes <- data.frame(
  sample = colnames(counts_raw),
  raw    = colSums(counts_raw),
  ftd    = colSums(counts_ftd),
  norm   = colSums(norm_counts)
) %>%
  pivot_longer(-sample, names_to = "stage", values_to = "reads")

ggplot(libsizes, aes(x = sample, y = reads / 1e6, fill = stage)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Library sizes (millions)", x = "Sample", y = "Reads (M)", fill = "Stage") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5.3 Boxplots and density

```{r boxplots}
for (nm in c("Raw" = "counts_raw", "Filtered" = "counts_ftd", "Normalized" = "norm_counts")) {
  mat <- get(nm)
  df <- long_expr(log2p1(mat), metadata)

  p <- ggplot(df, aes(x = sample, y = expr)) +
    geom_boxplot(outlier.size = 0.2) +
    labs(
      title = paste("Boxplot log2(count+1) -", nm),
      x = "Sample", y = "log2(count+1)"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p)
}
```

```{r density-plot}
df_den <- long_expr(log2p1(norm_counts), metadata)

ggplot(df_den, aes(x = expr, group = sample)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of log2(count+1) - Normalized", x = "log2(count+1)", y = "density") +
  theme_bw()
```

## 5.4 PCA and MDS

```{r pca-mds}
datasets <- list(
  Raw = counts_raw,
  Filtered = counts_ftd,
  Normalized = norm_counts
)
datasets_log <- lapply(datasets, log2p1)

for (nm in names(datasets_log)) {
  print(do_pca_plot(datasets_log[[nm]], metadata, sprintf("PCA - %s (log2+1)", nm)))
  print(do_mds_plot(datasets_log[[nm]], metadata, sprintf("MDS - %s (log2+1)", nm)))
}
```

## 5.5 PCA of normalized counts

```{r pca-normalized}
pca <- prcomp(t(log2p1(norm_counts)), center = TRUE, scale. = FALSE)

ev <- round(100 * (pca$sdev^2 / sum(pca$sdev^2))[1:2], 2)
pca_df <- transform(
  as.data.frame(pca$x[, 1:2]),
  sample    = rownames(pca$x),
  subject   = factor(metadata$subject),
  condition = factor(metadata$condition,
                     levels = c("Pre-Abx", "Pre-Abx+LPS", "Post-Abx", "Post-Abx+LPS"))
)

group_colors <- c(
  "Pre-Abx"      = "#804D80CC",
  "Pre-Abx+LPS"  = "#804D80CC",
  "Post-Abx"     = "#804D80CC",
  "Post-Abx+LPS" = "#804D80CC"
)

shape_map <- c(
  "Pre-Abx"      = 1,
  "Pre-Abx+LPS"  = 2,
  "Post-Abx"     = 19,
  "Post-Abx+LPS" = 17
)

ggplot(pca_df, aes(PC1, PC2, color = condition, shape = condition, label = subject)) +
  geom_point(size = 3) +
  geom_text(vjust = 2, size = 1) +
  scale_color_manual(values = group_colors, breaks = levels(pca_df$condition), name = NULL) +
  scale_shape_manual(values = shape_map, breaks = levels(pca_df$condition), name = NULL) +
  labs(
    title = "RNA-seq - PCA of Normalized Counts",
    x = sprintf("PC1 (%.2f%%)", ev[1]),
    y = sprintf("PC2 (%.2f%%)", ev[2])
  ) +
  theme_minimal(base_size = 6)
```

# 6. Differential expression overview

## 6.1 Volcano plots

```{r volcano-helpers}
alpha   <- 0.05
lfc_thr <- log2(1.5)

nm_pretty <- function(x) {
  x <- as.character(x)
  if (exists("label_map")) {
    lab <- unname(label_map[x])
    lab[is.na(lab) | lab == ""] <- x[is.na(lab) | lab == ""]
    return(lab)
  }
  x
}

ev_volcano <- function(df,
                       title,
                       select_labels = character(0),
                       label_col = "gene_id",
                       p_col = "padj",
                       fc_col = "log2FoldChange",
                       p_cut = alpha,
                       fc_cut = lfc_thr,
                       xlim_override = NULL) {
  stopifnot(all(c(label_col, p_col, fc_col) %in% names(df)))

  lab_plain <- as.character(df[[label_col]])
  pvals     <- as.numeric(df[[p_col]])
  lfcs      <- as.numeric(df[[fc_col]])

  ok   <- !is.na(pvals) & !is.na(lfcs)
  up_n <- sum(ok & pvals < p_cut & lfcs >=  fc_cut)
  dn_n <- sum(ok & pvals < p_cut & lfcs <= -fc_cut)
  cap_txt <- sprintf("DEG: Up = %d, Down = %d", up_n, dn_n)

  keyvals <- ifelse(lfcs < -fc_cut & pvals < p_cut, "royalblue",
                    ifelse(lfcs >  fc_cut & pvals < p_cut, "red2", "grey30"))
  keyvals[is.na(keyvals)] <- "grey30"
  names(keyvals)[keyvals == "red2"]      <- "Upregulated"
  names(keyvals)[keyvals == "royalblue"] <- "Downregulated"
  names(keyvals)[keyvals == "grey30"]    <- "Not significant"

  sel    <- unique(select_labels)
  is_sel <- lab_plain %in% sel

  lab_disp <- lab_plain
  if (any(is_sel)) lab_disp[is_sel] <- paste0("italic('", lab_plain[is_sel], "')")

  df2 <- df
  df2[[label_col]] <- lab_plain
  df2[[p_col]]     <- pvals
  df2[[fc_col]]    <- lfcs

  xlim_final <- if (!is.null(xlim_override)) xlim_override else c(-5, 5)

  EnhancedVolcano::EnhancedVolcano(
    toptable = df2,
    lab = lab_disp,
    selectLab = lab_disp[is_sel],
    parseLabels = TRUE,
    x = fc_col, y = p_col,
    pCutoff = p_cut, FCcutoff = fc_cut,
    xlim = xlim_final,
    colCustom = keyvals,
    drawConnectors = TRUE, widthConnectors = 0.5, colConnectors = "grey40",
    max.overlaps = Inf, boxedLabels = TRUE, colAlpha = 0.3,
    title = paste(title),
    subtitle = "Differential gene expression - FC cutoff: 1.5; p-value cutoff: 0.05",
    caption = cap_txt,
    xlab = bquote(Log[2] ~ "fold change"),
    ylab = bquote(-Log[10] ~ adjusted ~ italic(P)),
    titleLabSize = 8, axisLabSize = 8, legendLabSize = 7,
    subtitleLabSize = 7, captionLabSize = 7, pointSize = 2, labSize = 3,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    legendPosition = "top", border = "partial"
  )
}

label_map <- c(
  PostAbx_vs_PreAbx           = "Post-Abx vs Pre-Abx",
  PreAbxLPS_vs_PreAbx         = "Pre-Abx + LPS vs Pre-Abx",
  PostAbxLPS_vs_PreAbxLPS     = "Post-Abx + LPS vs Pre-Abx + LPS",
  PostAbxLPS_vs_PostAbx       = "Post-Abx + LPS vs Post-Abx",
  PostAbx_vs_PreAbxLPS        = "Post-Abx vs Pre-Abx + LPS"
)

manual_labels <- list(
  PostAbx_vs_PreAbx = c("CCL2","CCL20","EDN1","CXCL8","IL1A","NFKBIA","NFKB1",
                        "IL1B","NR4A1","TNF","CSK","SLC16A6","FOLR3","CRYBG1","RHOQ"),
  PreAbxLPS_vs_PreAbx = c("PTGS2","EDN1","CXCL2","CCL2","NFKB1","IL1A","TNFAIP3",
                          "NR4A1","PTGER2","IL1B","NLRP3","CXCL3","CXCL8","TNF",
                          "SOD2","RIG1","LTBR","TRAF2"),
  PostAbxLPS_vs_PostAbx = c("F8A2"),
  PostAbxLPS_vs_PreAbxLPS = c("TNFRSF12A","SLA2","MS4A6A","IL9R","VNN3P","PRR5L","LAIR1"),
  PostAbx_vs_PreAbxLPS = c("F8A2","TNFRSF12A","VNN3P","IL9R","CNR2")
)
```

```{r volcano-plots}
for (nm in names(DEG_res_list)) {
  df  <- DEG_res_list[[nm]]
  sel <- if (!is.null(manual_labels[[nm]])) manual_labels[[nm]] else character(0)

  xlim_use <- if (nm %in% c("PostAbxLPS_vs_PostAbx", "PostAbxLPS_vs_PreAbxLPS", "PostAbx_vs_PreAbxLPS")) {
    c(-1, 35)
  } else {
    c(-5, 5)
  }

  print(ev_volcano(df, nm_pretty(nm), select_labels = sel, xlim_override = xlim_use))
}
```

## 6.2 DEG counts barplot

```{r deg-counts-barplot}
ensure_regulation <- function(df) {
  if (!"regulation" %in% names(df)) {
    stopifnot(all(c("log2FoldChange", "padj") %in% names(df)))
    df <- df %>%
      mutate(
        regulation = case_when(
          padj < alpha & log2FoldChange >=  lfc_thr ~ "Up",
          padj < alpha & log2FoldChange <= -lfc_thr ~ "Down",
          TRUE ~ NA_character_
        )
      )
  }
  df %>% filter(!is.na(regulation))
}

count_up_down <- function(df, dataset_name) {
  df <- ensure_regulation(df)
  tibble(
    Dataset = dataset_name,
    Status  = df$regulation
  ) |>
    count(Dataset, Status, name = "Count")
}

count_data <- bind_rows(lapply(names(deg_list_sig), function(nm) {
  count_up_down(deg_list_sig[[nm]], nm)
}))

count_data <- count_data %>%
  mutate(
    Dataset_lbl = coalesce(label_map[Dataset], Dataset),
    Dataset_lbl = factor(Dataset_lbl, levels = rev(c(
      "Post-Abx vs Pre-Abx",
      "Pre-Abx + LPS vs Pre-Abx",
      "Post-Abx + LPS vs Pre-Abx + LPS",
      "Post-Abx + LPS vs Post-Abx",
      "Post-Abx vs Pre-Abx + LPS"
    ))),
    Status = factor(Status, levels = c("Down", "Up"))
  )

ggplot(count_data, aes(x = Dataset_lbl, y = Count, fill = Status)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.8) +
  geom_text(
    aes(label = Count),
    position = position_dodge(width = 0.7),
    vjust = 0.5, hjust = -0.1,
    size = 1.5,
    family = "Arial"
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c(Down = "#327ebaff", Up = "#da6868ff")) +
  labs(title = "Differentially Expressed Genes", x = NULL, y = "Count", fill = NULL) +
  theme_minimal(base_size = 7)
```

# 7. Functional enrichment

## 7.1 ORA helpers

```{r enrichment-helpers}
sym_to_entrez <- function(symbols) {
  sy <- unique(as.character(symbols))
  mp <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = sy,
    keytype = "SYMBOL",
    column = "ENTREZID",
    multiVals = "first"
  )
  out <- unname(mp[sy])
  out <- as.character(out)
  unique(out[!is.na(out) & nzchar(out)])
}

run_ora <- function(entrez_ids) {
  if (length(entrez_ids) == 0) return(list(GO = NULL, KEGG = NULL, REACT = NULL))
  list(
    GO = tryCatch(
      enrichGO(
        gene = entrez_ids, OrgDb = org.Hs.eg.db, keyType = "ENTREZID",
        ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH", readable = TRUE
      ),
      error = function(e) NULL
    ),
    KEGG = tryCatch(
      enrichKEGG(
        gene = entrez_ids, organism = "hsa",
        pvalueCutoff = 0.05, pAdjustMethod = "BH"
      ),
      error = function(e) NULL
    ),
    REACT = tryCatch(
      enrichPathway(
        gene = entrez_ids, organism = "human",
        pvalueCutoff = 0.05, pAdjustMethod = "BH", readable = TRUE
      ),
      error = function(e) NULL
    )
  )
}
```

## 7.2 Run ORA across contrasts

```{r run-ora}
ora_results <- list()

for (ct in names(DEG_res_list)) {
  df <- DEG_res_list[[ct]]

  sig_subsets <- list(
    All  = df |> filter(padj < alpha & abs(log2FoldChange) >= lfc_thr) |> pull(gene_id) |> unique(),
    Up   = df |> filter(padj < alpha &  log2FoldChange >=  lfc_thr)    |> pull(gene_id) |> unique(),
    Down = df |> filter(padj < alpha &  log2FoldChange <= -lfc_thr)    |> pull(gene_id) |> unique()
  )

  for (nm in names(sig_subsets)) {
    syms <- sig_subsets[[nm]]
    if (length(syms) == 0) next

    entrez <- sym_to_entrez(syms)
    if (length(entrez) == 0) next

    ora <- run_ora(entrez)

    if (is.null(ora_results[[ct]])) ora_results[[ct]] <- list()
    ora_results[[ct]][[nm]] <- list(
      symbols = syms,
      entrez  = entrez,
      GO      = ora$GO,
      KEGG    = ora$KEGG,
      REACT   = ora$REACT
    )
  }
}

names(ora_results)
```

## 7.3 Example GO dotplot

```{r go-dotplot}
.parse_ratio <- function(df) {
  if (!"GeneRatio" %in% names(df)) return(rep(NA_real_, nrow(df)))
  num <- suppressWarnings(as.numeric(sub("/.*$", "", df$GeneRatio)))
  den <- suppressWarnings(as.numeric(sub("^.*/", "", df$GeneRatio)))
  out <- num / den
  out[!is.finite(out)] <- NA_real_
  out
}

go_dotplot_df <- function(df,
                          include_desc,
                          regex = TRUE,
                          top_n = Inf,
                          title = NULL) {
  req <- c("Description", "p.adjust", "Count")
  stopifnot(all(req %in% names(df)))

  df <- df %>%
    mutate(
      GeneRatio_num = .parse_ratio(cur_data()),
      GeneRatio_num = ifelse(is.na(GeneRatio_num),
                             Count / max(Count, na.rm = TRUE),
                             GeneRatio_num)
    )

  if (regex) {
    pat <- paste0("(", paste(include_desc, collapse = ")|("), ")")
    dff <- df %>% filter(str_detect(Description, regex(pat, ignore_case = TRUE)))
  } else {
    dff <- df %>% filter(Description %in% include_desc)
  }

  if (!is.finite(top_n)) top_n <- nrow(dff)
  dff <- dff %>% slice_head(n = top_n)
  if (nrow(dff) == 0) stop("No GO terms matched.")

  dff <- dff %>% arrange(p.adjust, desc(GeneRatio_num))
  dff$Description <- factor(dff$Description, levels = rev(unique(dff$Description)))

  ggplot(dff, aes(x = GeneRatio_num, y = Description)) +
    geom_point(aes(size = Count, color = p.adjust, alpha = 0.4)) +
    scale_size(range = c(0.5, 3)) +
    scale_x_continuous(name = "Gene ratio") +
    scale_y_discrete(name = NULL) +
    scale_color_gradient(name = "padj", low = "red2", high = "royalblue", trans = "reverse") +
    guides(size = guide_legend(order = 1), color = guide_colorbar(order = 2)) +
    labs(title = title %||% "GO enrichment (selected terms)") +
    theme_bw(base_size = 6)
}

go_post_vs_pre <- ora_results[["PostAbx_vs_PreAbx"]][["All"]][["GO"]]@result

terms_exact <- c(
  "cellular response to lipopolysaccharide",
  "response to tumor necrosis factor",
  "chemotaxis",
  "granulocyte migration",
  "regulation of inflammatory response"
)

go_dotplot_df(
  df = go_post_vs_pre,
  include_desc = terms_exact,
  regex = FALSE,
  title = "DEGs - Post-Abx vs Pre-Abx (GO:BP)"
)
```

# 8. GSEA heatmap

```{r gsea-heatmap}
hallmark_file <- file.path(gsea_dir, "GENE_SYMBOLS_GSEA_INFLAMMATORY.txt")

hall_inf <- read.delim(hallmark_file, stringsAsFactors = FALSE)
hall_inf <- hall_inf$GENE_SYMBOLS

PostAbx_vs_PreAbx_sig <- deg_list_sig[["PostAbx_vs_PreAbx"]]

deg_hall_inf <- PostAbx_vs_PreAbx_sig %>%
  filter(gene_id %in% hall_inf) %>%
  pull(gene_id)

norm_filt <- norm_counts_anno %>%
  filter(gene_name %in% deg_hall_inf)

norm_filt_num <- norm_filt
rownames(norm_filt_num) <- norm_filt_num$gene_name
norm_filt_num <- norm_filt_num[, c(3:14)]
norm_filt_num <- as.matrix(norm_filt_num)

norm_filt_num_post <- norm_filt_num[, c(1, 3, 5, 7, 9, 11)]

anno_col_post <- data.frame(
  Group   = metadata$condition,
  Subject = metadata$subject
)

anno_col_post <- as.data.frame(anno_col_post[c(1, 3, 5, 7, 9, 11), ])
colnames(anno_col_post) <- c("Condition", "Subject")

anno_col_post$Condition <- factor(
  anno_col_post$Condition,
  levels = c("Pre-Abx", "Post-Abx")
)

anno_col_post$Group_plot <- factor(
  ifelse(anno_col_post$Condition == "Pre-Abx", "A Pre-Abx", "B Post-Abx"),
  levels = c("A Pre-Abx", "B Post-Abx")
)

anno_col <- HeatmapAnnotation(
  Group = anno_col_post$Group_plot,
  col = list(
    Group = c(
      "A Pre-Abx"  = "#804D8066",
      "B Post-Abx" = "#804D80cc"
    )
  ),
  annotation_name_gp = gpar(fontfamily = "Arial", fontsize = 6),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "",
      title_gp = gpar(fontfamily = "Arial", fontsize = 6),
      labels_gp = gpar(fontfamily = "Arial", fontsize = 6)
    )
  )
)

spl <- droplevels(anno_col_post$Group_plot)
n_slices <- nlevels(spl)

mat_z <- t(apply(norm_filt_num_post, 1, scale))

colnames(mat_z) <- c("S8", "S8", "S9", "S9", "S11", "S11")

color_z <- colorRamp2(
  seq(min(mat_z), max(mat_z), length = 3),
  c("royalblue2", "#EEEEEE", "firebrick2"),
  space = "RGB"
)

h_z <- Heatmap(
  mat_z,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  col = color_z,
  column_labels = colnames(mat_z),
  name = "Z-score",
  top_annotation = anno_col,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 5.5, fontfamily = "Arial"),
  column_names_gp = gpar(fontsize = 6, fontfamily = "Arial"),
  column_split = spl,
  column_title = rep("", n_slices),
  show_parent_dend_line = FALSE,
  show_column_names = FALSE,
  clustering_distance_rows = "manhattan",
  clustering_distance_columns = "manhattan",
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(
    legend_direction = "horizontal",
    color_bar = "continuous",
    legend_width = unit(5, "cm"),
    title_position = "lefttop",
    title_gp = gpar(fontfamily = "Arial", fontsize = 6, fontface = "bold"),
    labels_gp = gpar(fontfamily = "Arial", fontsize = 6)
  )
)

draw(
  h_z,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom",
  merge_legends = TRUE
)
```

# Session information

```{r session-info}
sessionInfo()
```
